Notes on Learning Linear State Space Models

### 2015-09-01 notes/notes-linear-system-learning.tex
Minor fiddles and fixes

\subsubsection{Factorising the Precision}

Using Givens rotations, we can efficiently factorise the precision eigenvector matrix as follows.
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix} \tvvec & \tvvecorth \end{bmatrix} &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix}\begin{bmatrix} \tvsign & \zmat \\ \zmat & \tvsignorth \end{bmatrix} \begin{bmatrix} \tvrow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tvnull \end{bmatrix}
\end{IEEEeqnarray}

@@

We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix}
  \lgtv\mcnew \\ \mhrot\mcnew
 \end{bmatrix}
 & = &
 \begin{bmatrix}
  \mhrot\lgtv\mhrot\tr \\ \mhrot\tr
 \end{bmatrix} \nonumber     ,
\end{IEEEeqnarray}
%
which is clearly its own inverse. Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \den(\lgtv\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtv|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
 & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv)} \times \frac{\den(\lgtv\mcnew, \lgtm) }{\den(\lgtv, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
\end{IEEEeqnarray}
%
The first term is simply a ratio of Kalman filter likelihoods.

There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$. For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
%
\begin{IEEEeqnarray}{rCl}
 P(S) & = & (\idmat - S)\inv(I+S)     .
\end{IEEEeqnarray}
%
To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
%
\begin{IEEEeqnarray}{rCl}
 s_{k,l} & \sim & \normaldist{0}{\sigma_s^2} \label{eq:skewsymmetric_proposal}     .
\end{IEEEeqnarray}
%
Use these to construct a skew-symmetric matrix $S$,
%
\begin{IEEEeqnarray}{rCl}
 S_{k,l} & = & \begin{cases}
                s_{k,l}  & k<l \\
                -s_{l,k} & k>l \\
                0        & k=l     ,
               \end{cases}
\end{IEEEeqnarray}
%
and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S\tr)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)=\mhrotppsl(\mhrot\mcnew)$, leading to a cancellation in the acceptance probability.

@@

The density of the proposal distribution for the new eigenvector is simply the reciprocal of the volume of the set, which is a hypersphere. Thus,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) &=& \frac{\gammafun\left(\frac{\lsd-\rk}{2}\right)}{2 \pi^{\half(\lsd-\rk)}}      .
\end{IEEEeqnarray}

@@

So,
%
\begin{IEEEeqnarray}{rCl}
 \mhar  &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
 \frac{ \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) }{ \ppslden{\eval{}}(\eval{\rk\mcnew}|\lgtv) \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) } \nonumber \\
 &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
 \frac{ \pi^{\half(\lsd-\rk)} \prod_{i=1}^{\rk}\eval{i}^{-1} }{ \gammafun\left(\frac{\lsd-\rk}{2}\right) } \times
 \frac{\eval{\rk+1}^{\lsd-\rk-1}\prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})}{\ppslden{\eval{}}(\eval{\rk+1}|\lgtv)} \nonumber      .
\end{IEEEeqnarray}

@@

>>>>

\subsubsection{Factorising the Covariance}

Using Givens rotations, we can efficiently factorise the covariance eigenvector matrix as follows.
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix} \tvvec & \tvvecorth \end{bmatrix} &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix}\begin{bmatrix} \tvsign & \zmat \\ \zmat & \tvsignorth \end{bmatrix} \begin{bmatrix} \tvrow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tvnull \end{bmatrix}
\end{IEEEeqnarray}

@@

We sample a rotation matrix $\mhrot$ from some proposal distribution with density $\mhrotppsl$ (w.r.t. Haar measure on orthogonal matrices) and then apply the following transformation,
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix}
  \lgtv\mcnew \\ \mhrot\mcnew
 \end{bmatrix}
 & = &
 \begin{bmatrix}
  \mhrot\lgtv\mhrot\tr \\ \mhrot\tr
 \end{bmatrix} \nonumber     ,
\end{IEEEeqnarray}
%
which is clearly its own inverse. Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \den(\lgtv\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtv|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
 & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv)} \times \frac{\den(\lgtv\mcnew, \lgtm) }{\den(\lgtv, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
\end{IEEEeqnarray}
%
The first term is simply a ratio of Kalman filter likelihoods.

There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$. For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
%
\begin{IEEEeqnarray}{rCl}
 P(S) & = & (\idmat - S)\inv(I+S)     .
\end{IEEEeqnarray}
%
The matrix exponential would also work. Note that the Jacobian of the Cayley transformation is derived in \cite{Olkin1953}, but this is not needed as it cancels out.

To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
%
\begin{IEEEeqnarray}{rCl}
 s_{k,l} & \sim & \normaldist{0}{\sigma_s^2} \label{eq:skewsymmetric_proposal}     .
\end{IEEEeqnarray}
%
Use these to construct a skew-symmetric matrix $S$,
%
\begin{IEEEeqnarray}{rCl}
 S_{k,l} & = & \begin{cases}
                s_{k,l}  & k<l \\
                -s_{l,k} & k>l \\
                0        & k=l     ,
               \end{cases}
\end{IEEEeqnarray}
%
and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S\tr)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)=\mhrotppsl(\mhrot\mcnew)$, leading to a cancellation in the acceptance probability.

@@

The density of the proposal distribution for the new eigenvector is simply the reciprocal of the volume of the set, which is a half of a hypersphere (since we have to resolve the eigenvector sign ambiguity). Thus,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) &=& \frac{\gammafun\left(\frac{\lsd-\rk}{2}\right)}{\pi^{\half(\lsd-\rk)}}      .
\end{IEEEeqnarray}

@@

So,
%
\begin{IEEEeqnarray}{rCl}
 \mhar  &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
 \frac{ \jacob(\lgtv,\eval{\rk\mcnew},\evec{\rk\mcnew} \to \lgtv\mcnew) }{ \ppslden{\eval{}}(\eval{\rk\mcnew}|\lgtv) \ppslden{\evec{}}(\evec{\rk\mcnew}|\lgtv) } \nonumber \\
 &=& \frac{ \den(\ls{1:\timax}|\lgtm, \lgtv\mcnew) \den(\lgtm,\lgtv\mcnew|\rk\mcnew) \den(\rk\mcnew) }{ \den(\ls{1:\timax}|\lgtm, \lgtv) \den(\lgtm,\lgtv|\rk) \den(\rk) } \times 
 \frac{ \pi^{\half(\lsd-\rk)} \prod_{i=1}^{\rk}\eval{i}^{-1} }{ 2 \: \gammafun\left(\frac{\lsd-\rk}{2}\right) } \times
 \frac{\eval{\rk+1}^{\lsd-\rk-1}\prod_{i=1}^{\rk}(\eval{i}-\eval{\rk+1})}{\ppslden{\eval{}}(\eval{\rk+1}|\lgtv)} \nonumber      .
\end{IEEEeqnarray}

@@

\section{Additional Notes}

I originally focussed on methods where the prior dependence between $\lgtm$ and $\lgtv$ was broken, and the two were sampled alternately. To do this efficiently (with matrix normal distribution for the transition matrix) requires using eigen-decompositions in a manner inspired by this paper: \cite{Stegle2011}.
